Modelling state-switching functional data with hidden Markov models

### 2020-11-05 Report/Sections/discussion.tex
Commit before adding t_tilde take 2

We demonstrate the usefulness of this framework using an ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a Northern Resident Killer Whale off the coast British Columbia, Canada. This gives a deeper understanding of the animal's tri-axial movement and thus its energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both Northern Resident Killer Whales and their prey \citep{Noren:2011}. This model is likely applicable to many other diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}. In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.

>>>>

We demonstrate the usefulness of this framework using an ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale off the coast British Columbia, Canada. This gives a deeper understanding of the animal's tri-axial movement and thus its energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both northern resident killer whales and their prey \citep{Noren:2011}. This model is likely applicable to many other diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}. In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.

### 2020-11-05 Report/Sections/simulation.tex
Commit before adding t_tilde take 2

%\section{Simulation Study}
We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,t^*}$ is set to a scalar rather than a three dimensional vector. We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information under that model.

@@

The two simulated dive types differ in that dives of type one are much shorter on average (20 seconds) than dives of type two (80 seconds). The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher for subdive state 2 than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data. All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of average subdive decoding accuracy, parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model does not estimate dive type as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a single Gamma distribution over the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and on average the value of $\hat \sigma$ is approximately 5 seconds less than the average sample standard deviation of dive duration, indicating significant model misspecification. See Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive types and subdive states for each model.

>>>>

%\section{Simulation Study}
We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known. The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity. In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,t^*}$ is set to a scalar rather than a three dimensional vector. We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT. Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times. To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors of a given model's parameter estimates with the standard errors estimated from the inverse of the observed Fisher information under that model.

@@

The two simulated dive types differ in that dives of type 1 are much shorter on average (20 seconds) than dives of type 2 (80 seconds). The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher for subdive state 2 than for subdive state 1. These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data. All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.

@@

Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of average subdive decoding accuracy, parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates. In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}). However, this model does not estimate dive type as it lacks any hierarchical structure. The CarHMM-DFT nonetheless fits a single Gamma distribution to the dive duration of all dives. The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and on average the value of $\hat \sigma$ is approximately 5 seconds less than the true sample standard deviation, indicating significant model misspecification. Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} visualize the simulated data set as well as the decoded dive types and subdive states for each model.
